---
title: "Introduction To R | Homework 5"
author:
  name: "Firstname Secondname"
  email: "my_email@lms.mrc.ac.uk"
date: "`r Sys.Date()`"
output: html_notebook
---

## Preliminaries

Make sure your working directory is set to the folder containing the course materials. You can do this with the user interface in `RStudio` (`Session > Set Working Directory`). Alternatively, you can get and set the working directory with commands.

```{r, eval=FALSE}
getwd()
```

```{r, eval=FALSE}
setwd("/this/is/a/location")  # Mac / Linux
setwd("C:\this\is\a\location")  # Windows
```

## Instructions

- Please fill in your name and email address in the header section at the top of the file.

- This document is an `R Notebook`. When you save, it will automatically render an HTML output with the result. You can upload either this raw document or the rendered HTML when submitting your work.

*Don't forget that you can insert a code block at the position of the cursor with `⌘+⌥+I` (Mac) or `CTRL+Alt+I` (Windows). These are run by pressing the `⏵` (run) button in the top right corner of each chunk.*

## Questions

We're going to be applying some of our learning with the `tidyverse` using data on the extent of Antarctic sea ice - this is a significant factor relating to breeding success in penguins!

The data are taken from the [NSIDC Sea Ice Index](http://nsidc.org/data/seaice_index), which collates daily measurements of the extent of sea ice (in millions of square kilometers). Here, we're working with the 5-day trailing averages to even out small daily fluctuations.

```{r}
ice = read_tsv("data/5_sea_ice.txt")
```

### 1

1) Exploring the dataset:

a) What makes this table tidy or untidy?

b) Pivot the table so that `year` is a column.

c) Use a `dplyr` function to add a column, `day_of_year` that we'll use for plotting.
(Hint: the `lubridate::make_date()` function can be used to construct a date from columns and `lubridate::yday()` will return the day of the year). 

### 2

2) Construct a `ggplot2` graph showing the extent of the sea ice on the `y` axis versus `day_of_year` on the `x`, colouring by `year`.

a) Display this with a `geom_line()`.

b) We rarely get to use any of `ggplot2`'s more exotic plotting methods. With this data, try out `coord_polar()` ...

### 3

3) Use `dplyr` functions, especially `group_by()` and `summarise()`, to answer the following:

a) Which `year`-`month` grouping has the lowest average sea ice extent?

b) Taking all years together, which `day_of_year` has the highest average sea ice extent?

c) For which day of the year is the largest difference between the highest and lowest extent value recorded?
