---
title: "Introduction To R | Homework 2"
author:
  name: "Firstname Secondname"
  email: "my_email@lms.mrc.ac.uk"
date: "`r Sys.Date()`"
output: html_notebook
---

## Preliminaries

Make sure your working directory is set to the folder containing the course materials. You can do this with the user interface in `RStudio` (`Session > Set Working Directory`). Alternatively, you can get and set the working directory with commands.

```{r, eval=FALSE}
getwd()
```

```{r, eval=FALSE}
setwd("/this/is/a/location")  # Mac / Linux
setwd("C:\this\is\a\location")  # Windows
```

## Instructions

- Work your way through the questions below. Some answers can be filled in as plain text but, for some, you will need to write some code!

- Please fill in your name and email address in the header section at the top of the file.

- This document is an `R Notebook`. When you save, it will automatically render an HTML output with the result. You can upload either this raw document or the rendered HTML when submitting your work.

*Don't forget that you can insert a code block at the position of the cursor with `⌘+⌥+I` (Mac) or `CTRL+Alt+I` (Windows). These are run by pressing the `⏵` (run) button in the top right corner of each chunk.*

## Questions

### 1

a) What is wrong with this code chunk?

```{r, eval=FALSE}
data = scan(data/1_scan.txt)
```

### 2

Your dog ate your homework! Unfortunately, your teacher doesn't believe you and you're forced to put it back together with sticky tape.

a) Can you correct this jumbled code?

```{r, eval=FALSE}
  "outputs/2_seq_depth.svg",
  width=8,
data = read.delim(

seq_depth = apply(data, 2, sum)
barplot(
  seq_depth,
dev.off()
  ylim=range(pretty(c(0, seq_depth))),
  main="Sequencing depth")

svg(
  col=c(rep("purple", 4), rep("green4", 4)),
  height=8)
  row.names=1)
data = as.matrix(data)
  "data/2_penguin_expression.tsv",
```

b) Describe what it does.

c) What does the `pretty()` function do here?

### 3

You have now imported the results of an Illumina sequencing experiment on two sets of penguins, 4 `Adelie`s and 4 `Gentoo`s.

a) How many genes are in the dataset?

b) Use `apply()` to calculate the row sums for the dataset.

c) For something as common as making a row sum, `R` has lots of built-in functions. Is there one that does this?

d) How many genes aren't expressed at all? (Hint ... you can `sum()` logical values)

e) Filter the matrix to keep only the genes that were detected.

### 5

As each sample in this dataset has been sequenced to a different total depth, we should normalise them so that the columns are comparable to one another.

a) Turn the `seq_depth` into a scaling factor by dividing it by its `min()`.

b) Apply this normalisation factor down the expression matrix.

### 6

Now that the samples are comparable, some collaborators have asked you to share plots of the expression of a few specific genes.

The following code will make a `stripchart()` for a given `gene_name`.

```{r, eval=FALSE}
gene_name = "PENG0000000007"

gene_data = data.frame(
  exprs=data[gene_name, ],
  group=c(rep("Adelie", 4), rep("Gentoo", 4)))

stripchart(
  exprs~group,
  data=gene_data,
  method="jitter",
  col=c("purple", "green4"),
  main=gene_name,
  ylab="Read count",
  xlim=c(0, 3),
  ylim=range(pretty(c(0, gene_data$exprs))),
  vertical=TRUE)
```

a) Wrap this code in a `for` loop that generates plots within the same PDF document. They want plots for their two favourite genes: `PENG0000000007` and  `PENG0000017646`.
