---
title: "Introduction To R | Homework 1"
author:
  name: "Firstname Secondname"
  email: "my_email@lms.mrc.ac.uk"
date: "`r Sys.Date()`"
output: html_notebook
---

## Preliminaries

Make sure your working directory is set to the main folder for the course. You can do this with the user interface in `RStudio` (`Session > Set Working Directory`). Alternatively, you can get and set the working directory with commands.

```{r, eval=FALSE}
getwd()
```

```{r, eval=FALSE}
setwd("this/is/a/location")
```

## Instructions

- Work your way through the questions below. Some answers can be filled in as plain text but, for some, you will need to write some code!

- Please fill in your name and email address in the header section at the top of the file.

- This document is an `R Notebook`. When you save, it will automatically render an HTML output with the result. You can upload either this raw document or the rendered HTML when submitting your work.

*Don't forget that you can insert a code block at the position of the cursor with `⌘+⌥+I` (Mac) or `CTRL+Alt+I` (Windows). These are run by pressing the `⏵` (run) button in the top right corner of each chunk.*

## Questions

### 1

Assuming we have two numerical vectors ...

```{r}
x = c(1, 3, 5, 7, 9)
y = c(2, 3, 5, 7, 11, 13)
```

... try to explain what happens in each case:

a) `x + 1`

b) `x + y`

c) `y[3]`

d) `y[-3]`

e) `y[x]`

f) `x + y`

### 2

a) What does the `dir()` function do?

b) What class of object is returned when you run `dir()`?

c) How long is the output for this function?

d) How might you reverse (sort backwards) the output of this function?

### 3

a) What is wrong with this code chunk?

```{r, eval=FALSE}
2019_penguin_data = read.delim("data/1_2019_penguin_data.tsv")
```

b) Correct the problem. Having successfully imported it, what class is the resulting object?

c) Compare your imported data to the original file (open it in a Text Editor or in Excel). What's different?

d) By consulting the help, can you work out what's going on? Why?

### 4

Your interest in penguins has grown significantly over time. Naturally, you're interested in looking over the [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/) dataset that is available for helping people to practice `R`!

a) Load the `data/1_palmerpenguins.csv` file.

b) Familiarise yourself with the dataset. Which columns might be better represented as factors?

c) Make any changes you think necessary.

### 5

a) What is the average `bill_length_mm` for the `Adelie` penguins?

b) How many `Gentoo` penguins were recorded in each year of the study?

c) Add a new column named `bill_ratio`, calculated as `bill_length_mm` divided by `bill_depth_mm`.

### 6

a) Make a `boxplot()` showing the `bill_ratio` for each `species`.

b) Which species has the highest `bill_ratio`?

b) Plot something else that interests you.

c) Consult the help for your chosen graph and make an aesthetic (title, axis names, colours, symbols, etc) change to it.
