---
title: "Introduction To R | Homework 3"
author:
  name: "Firstname Secondname"
  email: "my_email@lms.mrc.ac.uk"
date: "`r Sys.Date()`"
output: html_notebook
---

## Preliminaries

Make sure your working directory is set to the folder containing the course materials. You can do this with the user interface in `RStudio` (`Session > Set Working Directory`). Alternatively, you can get and set the working directory with commands.

```{r, eval=FALSE}
getwd()
```

```{r, eval=FALSE}
setwd("/this/is/a/location")  # Mac / Linux
setwd("C:\this\is\a\location")  # Windows
```

## Instructions

- Please fill in your name and email address in the header section at the top of the file.

- This document is an `R Notebook`. When you save, it will automatically render an HTML output with the result. You can upload either this raw document or the rendered HTML when submitting your work.

*Don't forget that you can insert a code block at the position of the cursor with `⌘+⌥+I` (Mac) or `CTRL+Alt+I` (Windows). These are run by pressing the `⏵` (run) button in the top right corner of each chunk.*

## Tasks

It's really useful to be able to re-create a plot that you might be given or to match up the aesthetics of your figures with those from someone else.

Using the provided object ...

```{r}
load("data/4_penguin_results.Rda")  # provides the `res` `DESeqResults` object
summary(res)
```

### 1

Let's practice using `ggplot2` by doing our best to recreate a figure that we've seen `DESeq2` produce previously. Make a data frame from `res` and see how closely you can re-create the plot below, which graphs the `log2FoldChange` column against `baseMean`.

```{r}
library(DESeq2)

plotMA(res, alpha=0.05)
```

Make use of the fantastic [reference guide](https://ggplot2.tidyverse.org/reference/index.html) as well as the the built-in help for the aesthetics and geometries. If in doubt, a few key words into Google will often point you in the right direction!

Consider:

- the non-linear `x` axis scale and its labels

- the `y` axis limits and label orientation

- point sizes and shapes

- a *manual* (hint!) scale for the colour aesthetic

- themes and appearance 

- reference lines

- point shape

### 2

2) Although they encourage us to weight significance more highly than effect size (the fold change), volcano plots remain confusingly popular as a means of displaying the results of differential expression analyses. Let's make one!

a) Make a volcano plot, colouring according to significance. For the uninitiated, a volcano plot is a graph of `-10*log10(padj)` on the `y` axis versus `log2FoldChange` on the `x`.

b) Add labels for the 5 genes with the most substantial negative `log2FoldChange` values. Experiment with using `geom_text()` and `geom_label()`. (Hint ... the `rownames` might benefit from being a column!)

c) `install.packages("ggrepel")` and try out a repel equivalent - `geom_text_repel()` or `geom_label_repel()`.
